Draft Article

### 2019-11-17 article.tex
wohlforth introduction edits and self theory edits

\begin{document}
\maketitle
% \newpage

@@

Not only was the OPM hack a public relations disaster, it had significant national security implications as well. US officials feared that the stolen data could compromise intelligence efforts by exposing undercover CIA officers working in embassies around the world, whose names would be suspiciously absent from OPM records. The information could also be used to identify American officials who might be susceptible to pressure, or worse, recruitment. FBI Director James Comey personally took questions about the incident. ``If you have my [application for security clearance],'' Comey said, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States. Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} With evidence that a foreign intelligence service had successfully obtained exactly that, one might expect that the United States would respond with a series of diplomatic measures intended to signal that highly invasive acts of espionage would not be tolerated from its international adversaries.

@@

The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack appears to fall within the bounds of a traditional intelligence operation, it will merit the diplomatic consequences of a traditional intelligence operation. US policymakers rarely acknowledge that this distinction exists---officially, all cyberattacks are illegal international behavior and will receive a strong response---but occasionally, in moments of candor, they will admit that cyber-espionage is different. A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footnote{In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets. See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence, not prosecution, is deemed to be the appropriate response. Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did. If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015}

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Aug 24 13:46:46 2019
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
Type & Incidents & Official responses & Response \% \\
  \hline
Defacement &   3 &   2 & 66.67 \\
  Data destruction &   8 &   5 & 62.50 \\
  Doxing &   6 &   3 & 50.00 \\
  Sabotage &  14 &   5 & 35.71 \\
  Denial of service &  17 &   6 & 35.29 \\
  Espionage & 231 &  44 & 19.05 \\
   \hline
\end{tabular}
\caption{Cyber-incident response rates by type of incident}
\label{response-pct}
\end{table}

In fact, cyber-espionage is routinely treated as a lesser category of offense by all nations, not just the United States. Of the hundreds of state-sponsored cyberattacks about which there is public knowledge, attacks classified as espionage are by the far most common---and the least likely to garner an official response from the victim government (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly. Of the 323 listed incidents, 38 were excluded from this calculation because they lacked an identifiable Victim, Sponsor, or Type. For more, see \cite{council_on_foreign_relations_new_2019}} The routine minimization of espionage is most easily observed through the plethora of modern cyber-incidents, but it has its roots in norms that are much older.

At first glance this observation might seem almost tautological. Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage. That formulation, however, begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage in cyberspace, what purpose that distinction servers, or whether it is prudent to continue making it. Certainly the US should respond proportionally to attacks on its systems, but what if the damage done to national security by espionage is especially significant? ``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'' said one anonymous senior intelligence official of the OPM hack. ``Clearly, it does.''\footcite{sanger_u.s._2015} As as far as the public knows, however, the size of operation did \emph{not} change the nature of the diplomatic response. Therefore, to understand why the Director of the CIA treats cyber-espionage as \enquote{fair game,} one first needs to understand why traditional espionage has historically that treatment as well.

@@

This raises a fascinating paradox: for espionage to be mutually permissible, both states must agree that it is so. But that requires \emph{both} states to determine they are getting more from practicing espionage against their adversary than their adversary is getting from practicing it against them. Otherwise, one would simply defect from that equilibrium and punish espionage at a level that imposes direct costs. Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it gets from espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary. I argue that the latter is true.

@@

Before continuing, it is necessary to properly define what I mean by espionage norms. Simply put, espionage norms dictate what constitutes espionage and how states are expected to respond when it is conducted against them.\footnote{There are, of course, other \enquote{norms} that involve espionage. The so-called \enquote{Moscow Rules} dictate how an agent or case officer is supposed to conduct themselves in hostile territory. In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} What consequences does a state impose when it uncovers an espionage operation run by another state? Espionage norms prescribe that the consequences remain minimal. Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them. In the rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again. In this article I present a theory for why those norms exist.

Prior literature has attempted to explain espionage norms using international law, but in the absence of any significant treaties governing peacetime espionage, scholars are often left without definitive answers.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage. In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments. See \cite{wright_essays_1962}} Since espionage is illegal everywhere, how can every state acknowledge that allies and adversaries alike are guaranteed to engage in it?\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is. If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero. This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} There is no legally consistent way to resolve that paradox, so scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact. Jared Beim determines that there while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so the prohibition on espionage merits further enforcement.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe. For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that his conclusion will have little impact on the practice of espionage, where, at best, \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A. John Radsan employs a more philosophical approach. Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage all reach the same conclusion: it would be almost impossible to argue that espionage is legal, but it will likely continue to be downplayed. Beyond simple inertia, however, this literature cannot explain why espionage is tolerated.

Another branch of scholarship sidesteps the legal question in favor of a \enquote{functional theory} of IR. This is closer to my work, as it attempts to explain \emph{why} states tolerate espionage rather than creating \emph{post hoc} justifications for it. A functional approach, for instance, might highlight the role of espionage in treaty verification, the way Christopher Baker does when he argues that \enquote{the advantages that espionage offers over legally-binding verification and assurance regimes tip the scales in favor of functional cooperation.}\footcite{baker_tolerance_2004} Baker correctly notes that the additional assurance provided by espionage enables states to enter into riskier agreements that they would otherwise have avoided. His analysis is limited, however, to arenas in which espionage may facilitate various forms of interstate cooperation. It doesn't resolve the gap between the absolute illegality of espionage in domestic law and its widespread practice internationally. Responding to Baker, Radsan notes that one-off deals and treaties \enquote{are a very slow and indirect way to achieve international consensus on the legality of espionage. The gap is still there.}\footcite[p.~607]{radsan_unresolved_2007} This article expands on Baker's basic premise---that there are situations in which espionage is mutually advantageous for both parties---and universalizes it to explain why the entire institution of espionage is mutually advantageous. Espionage norms provide the glue that keep the institution in place, despite its legal ambiguity.

@@

\section{Theory}
One of the most prevalent clich\'es in spy literature and journalism is the so-called \enquote{rules of the game.} Media-friendly lists of rules like \enquote{You are never alone} and \enquote{Don't trust anyone} make the spy world seem insular, mysterious, and impenetrable.\footcite{myre_moscow_2019} They hint at a larger set of processes whose motivations the average citizen, not given access to this privileged information, could not possibly begin to understand. In reality, while the \emph{operations} of a modern intelligence service are incredibly opaque, the geopolitical conditions in which it operates are not. Every aspect of a nation-state's interactions with another ought to be subject to the same critical analysis. Scholars can use rationalist principles to analyze everything from trade policy to the space race; espionage, which by definition involves deeply invasive interactions with another state, ought to be no different.

Espionage certainly seems like something that states should be doing more to discourage. Then results of intelligence operations can have devastating, even existential effects on the targeted nation's security. One of the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been relentless. A US Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} ``Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,'' the report read, ``all with little to no adverse consequences to the thieves.''\footcite[p.~6]{bayer_cybersecurity_2019} The Navy's forecasts were dire: ``If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.''\footcite[p.~5]{bayer_cybersecurity_2019} The Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, the possibilities are alarming and the consequences for remain minimal.

% first it will lay out how espionage is considered an occupational hazard in the business of running a nation-state, one that is counteracted but never discourages.

% next I will discuss the theoretical justifications for why this might be the case. security seeking states and types of greedy states derive a reational benefit from allowing espionage to continue.

% finally I will look at historical moments in which new technologies were deployed for espionage-like purporses, and demonstrate how the norms governing the use of these technologies always evolved to permit espionage.

>>>>

\begin{document}
\maketitle

@@

Not only was the OPM hack a public relations disaster, it had significant national security implications as well. US officials feared that the stolen data could compromise intelligence efforts by exposing undercover CIA officers working in embassies around the world, whose names would be suspiciously absent from OPM records. The information could also be used to identify American officials who might be susceptible to pressure, or worse, recruitment. The backlash was so intense that FBI Director James Comey was forced to take questions about the incident. ``If you have my [application for security clearance],'' Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States. Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} With evidence that a foreign intelligence service had successfully obtained exactly that, one might expect the United States to respond with a series of diplomatic measures intended to signal that highly invasive acts of espionage would not be tolerated from its international adversaries.

@@

The US government's tepid response to the OPM hack illustrates a crucial fact about the relationship between cybersecurity and espionage: when a cyberattack appears to fall within the bounds of a traditional intelligence operation, it will merit the diplomatic consequences of a traditional intelligence operation. US policymakers rarely acknowledge that this distinction exists---officially, cyberattacks are illegal international behavior and will receive a strong response---but occasionally, in moments of candor, policymakers will admit that cyber-espionage is different. A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footnote{In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets. See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence, not prosecution, is deemed to be the appropriate response. Director of the CIA John Brennan described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did. If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015}

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Aug 24 13:46:46 2019
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
Type & Incidents & Official responses & Response \% \\
  \hline
Defacement &   3 &   2 & 66.67 \\
  Data destruction &   8 &   5 & 62.50 \\
  Doxing &   6 &   3 & 50.00 \\
  Sabotage &  14 &   5 & 35.71 \\
  Denial of service &  17 &   6 & 35.29 \\
  Espionage & 231 &  44 & 19.05 \\
   \hline
\end{tabular}
\caption{Global cyber-incident response rates}
\label{response-pct}
\end{table}

In fact, cyber-espionage is routinely treated as a lesser category of offense by all nations, not just the United States. Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by the far most common---and the least likely to garner an official response from the victim government (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly. Of the 323 listed incidents, 38 were excluded from this calculation because they lacked an identifiable Victim, Sponsor, or Type. For more, see \cite{council_on_foreign_relations_new_2019}} The routine minimization of espionage is most easily observed through the plethora of modern cyber-incidents, but it has its roots in norms that are much older.

At first glance this observation might seem almost tautological. Of course cyber-espionage is treated like traditional espionage, one could argue---they're both espionage. That formulation, however, begs the question; it doesn't explain how states determine what constitutes \enquote{traditional} espionage in cyberspace, what purpose that distinction servers, or whether it is prudent to continue making it. Certainly the US should respond proportionally to attacks on its systems, but what if the damage done to national security by espionage is especially significant? ``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'' said one anonymous senior intelligence official of the OPM hack. ``Clearly, it does.''\footcite{sanger_u.s._2015} As as far as the public knows, however, the size of operation did \emph{not} change the nature of the diplomatic response. Therefore, to understand why the Director of the CIA treats cyber-espionage as \enquote{fair game,} one first needs to understand why traditional espionage has historically received that treatment as well.

@@

This hypothesis raises a fascinating paradox: for espionage to be mutually permissible, both states must agree to treat it that way, but, in theory, that requires \emph{both} states to determine they are getting more from practicing espionage against their adversary than their adversary is getting from practicing it against them. Otherwise, the state with less to gain would simply defect from that equilibrium and punish espionage at a level that imposes direct costs. Since this generally doesn't happen between major world powers, one of two things must be true: either a state is miscalculating the relative gains it gets from espionage, or the mutual tolerance of espionage has benefits that are sufficiently important to maintain even if one state gains less from it than its adversary. I argue the latter.

@@

Before continuing, it is necessary to properly define what I mean by espionage norms. Simply put, espionage norms dictate what constitutes espionage and how states are expected to respond when it is conducted against them.\footnote{There are, of course, other \enquote{norms} that involve espionage. The so-called \enquote{Moscow Rules} dictate how an agent or case officer is supposed to conduct themselves in hostile territory. In order to avoid constantly specifying, espionage norms are more strictly defined to include only the ones that affect IR.} What consequences does a state impose when it uncovers an intelligence operation run by another state? Espionage norms prescribe that the consequences remain minimal. Rarely does the response amount to more than a pat denunciation of the activity and implementation of the necessary countermeasures to end it---the consequences for espionage are borne by the agents who commit it, not the state that sent them. In the rare cases when additional costs \emph{are} imposed on a state for conducting a traditional espionage operation, they are not costs that would discourage the aggressor from attempting such an operation again. In this article I present a theory for why those norms exist.

Prior literature has attempted to explain espionage norms using international law, but in the absence of any significant treaties governing peacetime espionage, scholars are often left without definitive answers.\footnote{Because it is universally agreed that international law does not deal with peacetime espionage directly, many journal articles that deal with this question frame their work almost Socratically, devoting two sections to cases for and against the legality of espionage. In fact, the first major work on law and peacetime espionage is a series of commentaries collected in a single publication that respond to each other's arguments. See \cite{wright_essays_1962}} Since espionage is universally illegal, how can every state acknowledge that allies and adversaries alike are guaranteed to engage in it?\footnote{Further complicating matters, espionage is the rare crime that might not be a crime depending on who the victim is. If an American government employee walks out of The Pentagon and hands classified documents to the KGB, they are at risk prosecuted as a traitor; if an American case officer hands classified KGB documents over to the CIA, they will be lauded as a hero. This example, in various forms, is frequently used to illustrate the impossibility of creating a consistent legal doctrine for espionage.} As a result, scholars have typically treated espionage as \emph{sui generis}---starting from the assumption that espionage norms exist in a class of their own, then attempting to fit a legal doctrine around that fact. Jared Beim determines that while peacetime espionage is often tolerated, it clearly constitutes an \enquote{intervention} as defined by the International Court of Justice, so the prohibition on espionage merits further enforcement.\footnote{Beim also suggests that a weaker country could seek justice for an espionage operation by appealing to international organizations like the United Nations or the Council of Europe. For more, see \cite{beim_enforcing_2018}} Beim acknowledges, however, that his conclusion will have little impact on the practice of espionage, where, at best, \enquote{it is perhaps possible to stop the more egregious violations.}\footcite[p.~672]{beim_enforcing_2018} A. John Radsan employs a more philosophical approach. Suggesting that we should resist the \enquote{Hegelian impulse} to resolve the tension between espionage and international law, Radsan concludes that \enquote{beyond any international consensus, countries will continue to perform espionage to serve their national interests \textelp{} International law does not change the reality of espionage.} Legal scholars of espionage all reach the same conclusion: it would be almost impossible to argue that espionage is legal, but no one really seems to care. Taken to its logical endpoint, this branch of scholarship would simply argue that the espionage norms we observe today result from simple inertia.

Another branch of scholarship sidesteps the legal question in favor of a \enquote{functional theory} of IR. This approach is more informative because it attempts to explain why states tolerate espionage rather than creating \emph{post hoc} justifications for it. A functional approach, for instance, might highlight the role of espionage in treaty verification, the way Christopher Baker does when he argues \enquote{the advantages that espionage offers over legally-binding verification and assurance regimes tip the scales in favor of functional cooperation.}\footcite{baker_tolerance_2004} Baker correctly notes that the additional assurance provided by espionage enables states to enter into riskier agreements that they would otherwise have avoided. His analysis is limited, however, by its focus on enabling specific policy outcomes, like treaties. It doesn't resolve the gap between the absolute illegality of espionage in domestic law and its widespread practice internationally. Responding to Baker, Radsan notes that one-off deals and treaties \enquote{are a very slow and indirect way to achieve international consensus on the legality of espionage. The gap is still there.}\footcite[p.~607]{radsan_unresolved_2007} This article expands on Baker's basic premise---that there are situations in which espionage is mutually advantageous for both parties---and universalizes it to explain why the entire institution of espionage is mutually advantageous.

@@

\section{Theory}
\subsection{Establishing the puzzle}
One of the most prevalent clich\'es in spy literature and journalism is the so-called \enquote{rules of the game.} The supposed rules of the spy world world---\enquote{You are never alone}, \enquote{Don't trust anyone}---make the spy world seem insular, mysterious, and impenetrable.\footcite{myre_moscow_2019} They hint at a larger set of processes whose motivations the average citizen, not given access to this privileged information, could not possibly begin to understand. In reality, while the \emph{operation} of a modern intelligence service is incredibly opaque, the geopolitical situation in which one operates is not. Every aspect of a nation-state's interactions with another ought to be subject to the same critical analysis. Scholars can use rationalist principles to analyze everything from trade policy to the space race; espionage, which by definition involves deeply invasive interactions with another state, ought to be no different.

% Certainly there are factors that the go into diplomatic decisions that the general public will likely never know. There are few reasons to believe, however, that this is more true for intelligence-related diplomatic matters than for generic ones. Every diplomatic decision is made with the backing of classified intelligence.

Using traditional methods of analysis, espionage certainly seems like something that states should be doing more to discourage. The results of intelligence operations can have devastating, even existential effects on the targeted nation's security. Arguably the most traditional form of espionage is stealing military secrets from a foreign power, and in that arena China has been relentless. A United States Navy report from March 2019 declared Chinese intelligence operations so extensive that they had substantively altered the balance of power between the two states.\footcite{lubold_navy_2019} ``Long-term, US future military advantage is being diminished by years of IP exfiltration from the DoD, DoN, and DIB,'' the report read, ``all with little to no adverse consequences to the thieves.''\footcite[p.~6]{bayer_cybersecurity_2019} The forecasts were dire: ``If the current trend continues unimpeded, the US will soon lose its status as the dominant global economic power.''\footcite[p.~5]{bayer_cybersecurity_2019} The US Navy is not a disinterested party, of course, and there are reasons to take its conclusions with a grain of salt; modern military technology is remarkably complex and China cannot simply steal its way to military dominance.\footnote{For a good discussion of these limitations, see \cite{gilli_why_2019}} Nonetheless, espionage presents alarming opportunities for rival states and the consequences for taking advantage of them remain minimal.

Espionage has potentially disastrous consequences---and steps that might actually curb it are out of the question. The system of repercussions is set up so that the structure of civilian intelligence services is never fundamentally threatened. When an especially damaging espionage operation is uncovered, the traditional punishment is to declare certain embassy officials \emph{persona non grata}, which often results in \enquote{tit-for-tat} diplomatic expulsions. These expulsions make it more difficult for the affected state to conduct intelligence operations in the short-term, but they don't fundamentally punish the state for having attempted espionage in the first place.

The name \enquote{tit-for-tat} is taken from a highly-successful strategy for the prisoner's dilemma wherein the player simply mimics the choice that that their opponent made in the last round. By employing the tit-for-tat strategy, the player never gets taken advantage of for more than one turn and can always return to cooperation if their opponent unilaterally signals a willingness to do the same. Diplomatic expulsions play out in exactly the same fashion: when one country expels the diplomats of another, the other country will often respond in kind. Each side will continue to expel each other's diplomats until one side signals that it would like for the reciprocal PNG-ing of diplomats to stop. Then, both sides return to a cooperative state, each preserving some of their intelligence operations. Because neither side can gain an a decisive advantage, the back-and-forth punishment never ends in a mutual defection.

\subsection{Espionage norms as a discrete series of choices}
The key feature of \enquote{tit-for-tat} diplomatic expulsions is that the scope of the punishment is structurally incapable of discouraging a future, identical operation. Losing intelligence officers in an embassy temporarily makes it more difficult to conduct espionage, but it doesn't make a state less likely to try; the worst that can happen is a net-neutral result---going back to \enquote{square one,} so to speak. By contrast, were it more common to respond to espionage by imposing financial sanctions or even inflicting a military cost, a state would have to consider the possibility that attempting espionage could adversely impact other aspects of their security, leaving it worse off than when it started. These are the sorts of calculations that states need to make in other high-stakes arenas---pursuing a nuclear weapons program, for instance, could lead to severe economic retaliation, as could protectionist trade policies---but engaging in espionage has very little potential for negative externalities because espionage norms take the possibility of a severe response off the table. In this section, I present what could happen if they didn't.

% https://tex.stackexchange.com/questions/249480/creating-a-payoff-matrix-using-latex-tabular-environment
\begin{table}[ht]
\centering
\setlength{\extrarowheight}{2pt}
\small
\begin{tabular}{cc|c|c|}
  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{Country 2}\\
  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
  \multirow{3}*{Country 1}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only the USSR continues spying \\ ~} \\\cline{3-4}
  & Punish Espionage & \makecell{~\\ Only the US continues spying \\~} & \makecell{~\\ No intelligence either direction \\~} \\\cline{3-4}
\end{tabular}
\caption{Espionage consequences decision matrix}
\label{espionage-matrix}
\end{table}

% Possibly swap this with the previous paragraph

To describe how states cooperate to preserve the liminal legal status of espionage, I propose an iterated decision-making model. My model changes the parameters so that the decision to cooperate or defect is based not in the amount of punishment received, but rather the type. Each time a state is forced to respond to a successful intelligence operation by a rival---such as the reveal of a high-level spy or a major data breach---it can either limit its response to one which is traditionally associated with espionage, or it can impose unprecedented costs that might discourage such an operation in the future. Table \ref{espionage-matrix} describes a single instance of this choice. In the status quo, State 1 signals its interest in cooperating by limiting its retaliation in response to espionage, no matter how damaging the operation is to their national security. State 2 responds by doing the same. Each iteration is a chance to play the game with fresh information; each espionage case is an opportunity to cooperate, reaffirming the norm that states always respond to espionage in a specific way, or defect, adding an additional cost to the attempt.

Through this model two important things become clear. First, even though either state has the opportunity to defect by imposing harsher punishments in response to espionage, its gains would be short-lived, because the other state is guaranteed to immediately do the same---tit-for-tat. Espionage consequences are an iterated game that both parties are playing the exact same way, so, in the long run, the only two possible outcomes are mutual defection or mutual cooperation. In the realm of traditional espionage, we've only ever seen mutual cooperation, but the critical observation is that any state has the power to change that by defecting. Even if the defector's rival would have preferred to maintain the status quo, the rival has no option other than to defect as well or else be left at a senseless disadvantage. The relatively open flow of information is only possible with a universal buy-in.

Second, while there is unquestionably a cultural norm informing the response to intelligence operations, that norm is doing a lot of work to maintain the status quo. So much work, in fact, that tradition could not possibly explain the entire phenomenon. Stripped of its mystique, the benefits of allowing intelligence operations to continue are, like any other state decision, a simple cost-benefit analysis; states decide that it best serves their interests to remain in the top left corner, rather than the bottom right. We can expect that there is \emph{some} rational explanation that political leaders employ to justify it, where they weigh the benefits of preserving their ability to spy against the leaks that inevitably result from espionage conducted against them.

One possible explanation is that some states are just miscalculating. Though I have not found any scholarly works making this argument, the basic premise is persuasive. Every policymaker knows exactly how crucial espionage is to their job. With years of experience making difficult policy calls informed by intelligence, it is certainly much easier to imagine how one's decisions will be made worse without espionage than it is to imagine how one's enemies' decisions will be made better. It follows that there would be a strong psychological bias towards attempting to discover as much information as possible. While this is likely true to some extent, \enquote{everyone else is getting it wrong} is not an explanation that typically suffices in political science, and it will not suffice here.

The surprisingly straightforward conclusion is that, if only two worlds are possible---one with widespread espionage and one without---then all the relevant nations must be concluding that the world without espionage is incredibly bad. So bad that even if one nation correctly concludes that its rival gains more from a world with a free flow of information than it does, the alternative, a world with significantly harsher punishments for espionage and a staunched flow of intelligence, still leaves that nation worse off. That calculation has a name: defensive realism.

\subsection{Defensive realism}
To explain why states might be willing to accept relative losses in security, I employ Charles Glaser's \emph{Rational Theory of International Politics}. Glaser describes his theory as integrating defensive realism with neoclassical realism.\footnote{With apologies to Mr. Glaser, I am going to refer to it as defensive realism, because they both put the security dilemma at the center of their explanation for international competition, which is the key observation for me. He would have you know that defensive realism ``can be viewed as a way station along the route to the full theory'' that he develops, which is ``significantly more general and complete than is defensive realism.'' (p. 13-14)} Glaser begins from the core assumption that the reason for competition between security-seeking states is insecurity. Consider a theoretical security seeker interacting with its adversary, which it believes to be a security-seeking state as well. The first state feels insecure, so it builds up arms to increase its security. When the rival state, seeing that its adversary is building up arms, now feels less secure and builds up its own arsenal in response, and a classic security dilemma ensues. Even though both states were just security seekers, they're now both much less secure.

% If both states are genuine security-seekers, then this situation is clearly sub-optimal, because now both states are heavily militarized and comparatively much less secure.

Defensive realism introduces a counterintuitive observation: the key to interrupting the security dilemma is to make the adversary feel more secure, because it is the adversary's response to securitization that ultimately makes the first state insecure. Therefore, defensive realism posits that it is logical for a security seeker to take actions that increase not just its own security, but its \emph{adversary's} security as well.\footcite[p.~7]{glaser_rational_2010} These actions would signal to the adversary that the state is a security seeker, hoping to reach a mutually agreeable outcome. If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond in kind and resolve the security dilemma for both of them.

The problem for a defensive realist is that unilaterally taking steps to increase the adversary's security is obviously quite risky. The state could be wrong that its adversary is a security seeker, in which case its optimistic overtures of cooperation will have exposed it to greater risk of attack. When states are unwilling to make those signaling measures because they fear compromising their own security, then a security dilemma results. This is where intelligence comes into play.

Prior to making any decisions, states asses the security situation and determine which of their options are most likely to increase their own security. Glaser identifies two types of variables that mediate the magnitude of a security dilemma. The first are material variables, such as offense-defense balance and offense-defense distinguishability. Espionage plays a significant role in properly assessing material variables; during the Cold War, the United States placed a premium on generating quality estimates of how many bombers, missiles, and bases the Soviet Union had deployed. By knowing which of its adversary's capabilities are offensive, a state can better target its efforts towards reducing them.

In addition to material variables, Glaser notes that information variables play an important and comparatively under-studied role in the security dilemma. Better information about its adversary's motives can give the state reason to trust that their risky signaling measures might succeed---or fail. Even if the state is uncertain, espionage can provide some information about its adversary's intent, which, in turn, can moderate the security dilemma enough to make cooperation possible. In the best case-scenario, the security dilemma is simply eliminated if both states are confident that the other is a security seeker. This is the optimal outcome for two genuine security-seekers, and it is made much more likely from the intelligence that espionage can provide.

Consider a situation in which one state is a genuine security-seeker and so is its adversary. The state conducts espionage on its adversary, and the resulting intelligence increases its internal estimate of how likely the adversary is to be a security-seeker. As a result, the state decides that some form of arms control is no longer too risky to attempt in order to signal its benign intentions. Because the state receiving the signal is a security-seeker, not only will its own intelligence make it more likely to respond in kind, but it also benefited from the first state's intelligence operations, because that led to the initial exchange of signals. The two-way flow of information has a measurable impact on each state's decisionmaking, facilitating the optimal outcome for both parties.

\subsection{Introducing the empirical examples}
% I use defensive realism to formulate a theoretical foundation to explain the existence of espionage norms generally, and the Soviet Union's extension of espionage norms to spy satellites in particular. Between two security-seeking states, both states will be less secure in the event of an arms race, so preventing one is a mutually-agreeable outcome. Resolving the security dilemma requires that the states signal their benign intentions, which states are more likely to do if they have good intelligence---intelligence that informs them about their adversary's capabilities and intent. Therefore, cooperating to preserve the practice of espionage as a semi-legitimate state institution makes rational sense, both because of the benefits espionage offers the state doing the spying, and the benefits the same state derives \emph{from being spied upon}.
